---
title: "Final"
author: Molly Bratton and Audrey Moyer
output:
  html_document: default
  pdf_document: default
---

```{r, setup, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Reading in datasets

```{r}
library(ggplot2)
library(tidyverse)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(maps)
library(plotly)
library(usdata)
library(mapproj)
library(ggthemes)
library(leaflet)
```

```{r}
states <- map_data("state")

#reading in the datasets
public_schools <- read_csv("https://data-nces.opendata.arcgis.com/api/download/v1/items/6a4fa1b0434e4688b5d60c2e5c1dcaaa/csv?layers=0")

neighborhood_poverty <- read_csv("https://data-nces.opendata.arcgis.com/api/download/v1/items/10bd90462ec14a53bab1827bc8f533c9/csv?layers=1")
```
# Data wrangling

```{r}
#need to join the datasets - they are lined up in the same order (checked by school name and NCESSCH)

schools_poverty <- public_schools %>%
  left_join(neighborhood_poverty, by = "NCESSCH")
```

```{r}
#changing variable names to be actually understandable
schools_poverty <- schools_poverty %>%
  group_by(STABR) %>%
  mutate(
    "state" = tolower(abbr2state(STABR))
      )
```

```{r}
#filtering out schools w no students

schools_poverty <- schools_poverty %>%
  filter(TOTAL > 0) %>%
  filter(TOTFRL >= 0) %>%
  mutate(
    "Ratio_of_Free_Lunches" = (TOTFRL)/TOTAL
) %>%
  filter(Ratio_of_Free_Lunches <= 1) # need to remove schools that have a ratio 1> bc that doesn't make any sense
```


```{r}
s_p_names <- schools_poverty %>%
  group_by(state) %>%
  summarise(
    "reduced/free_lunches" = mean(TOTFRL, na.rm = TRUE),
    "Income_Poverty_Ratio" = mean(IPR_EST, na.rm = TRUE),
    "Student_Teacher_Ratio" = mean(STUTERATIO, na.rm = TRUE),
    "Black_Student_Ratio" = mean(BL/TOTAL, na.rm = TRUE),
    "Male_To_Female_Ratio" = mean((TOTMENROL+1)/(TOTFENROL+1), na.rm = TRUE),
    "Ratio_of_Free_Lunches" = mean(((TOTFRL)/(TOTAL)), na.rm = TRUE),
    "lon" = LON,
    "lat" = LAT
)
```

# Make intital maps

```{r}
ggplot(data = s_p_names) + 
  geom_map(
    aes(map_id = state, fill = Ratio_of_Free_Lunches), map = states) +
  scale_color_viridis_c(aesthetics = "fill") +
  expand_limits(x = states$long, y = states$lat) +
  coord_map() +
  theme_map() +
  theme(legend.position = "bottom") +
  labs(title = "Average ratio of students getting free/reduced lunch to the total number of students")
```



```{r}
MN_schools <- schools_poverty %>%
  filter(STABR== "MN")

mn_map_data <- map_data("county", region = "minnesota")

ggplot(mn_map_data, aes(x=long, y=lat, group=group)) +  
  geom_polygon(color="black", fill="white") + 
  coord_fixed() +
  theme_map() +
  geom_point(data = MN_schools, aes(x = LON, y = LAT, group = FALSE, colour = IPR_EST, alpha = 0.5)) +
  scale_color_viridis_b() +
  labs(title = "Average IPR for public schools in MN")
```


```{r}
schools_poverty <- schools_poverty %>%
  filter(STUTERATIO > 2.785) %>%
  filter(STUTERATIO < 26.265)
ggplot(schools_poverty, aes(x = STUTERATIO, y = IPR_EST)) +
      geom_point(alpha = 0.25) +
      geom_smooth(method = "lm", se = FALSE) 
```

```{r}
IQR(schools_poverty$STUTERATIO, na.rm = TRUE)

quantile(schools_poverty$STUTERATIO, probs = c(0,0.25,0.5,0.75,1), na.rm = TRUE)

#outliers below 2.785 and above 26.265 using 1.5IQR method
```


# Our Research Questions

\textbf{1.} Does a higher school neighborhood poverty estimate correlate with a higher or lower free lunch program at schools?

```{r}
ggplot(mn_map_data, aes(x=long, y=lat, group=group)) +  
  geom_polygon(color="black", fill="white") + 
  coord_fixed() +
  theme_map() +
  geom_point(data = MN_schools, aes(x = LON, y = LAT, group = FALSE, colour = TOTFRL, alpha = 0.5)) +
  scale_color_viridis_b()
```


\textbf{2.} Are more densely populated areas more likely to have a higher or lower teacher to student ratio?

```{r}
ggplot(mn_map_data, aes(x=long, y=lat, group=group)) +  
  geom_polygon(color="black", fill="white") + 
  coord_fixed() +
  theme_map() +
  geom_point(data = MN_schools, aes(x = LON, y = LAT, group = FALSE, colour = TOTAL, alpha = 0.5)) +
  scale_color_viridis_b()
```


```{r}
ggplot(mn_map_data, aes(x=long, y=lat, group=group)) +  
  geom_polygon(color="black", fill="white") + 
  coord_fixed() +
  theme_map() +
  geom_point(data = MN_schools, aes(x = LON, y = LAT, group = FALSE, colour = STUTERATIO, alpha = 0.5)) +
  scale_color_viridis_b() +
  labs(title = "Student to Teacher Ratio in Minnesota")
```


\textbf{3.} Does the school neighborhood poverty estimate correlate to the percentage/ratio of black students of a school (or non-white students)? 

```{r}
ggplot(states, aes(x=long, y=lat, group=group)) +  
  geom_polygon(color="navy", fill="lightblue") + 
  coord_fixed() +
  theme_map() +
  geom_point(data = s_p_names, aes(x = lon, y = lat, colour = Black_Student_Ratio, group = FALSE), shape = ".") #make point size smaller
#could filter data by low, medium, and high, and make three different maps
```


\textbf{4.} Does a higher student to teacher ratio correlate with higher or lower free lunch programs at schools?

```{r}
ggplot(schools_poverty, aes(x = STUTERATIO, y = TOTFRL)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Scatterplot of Student Teacher Ratio against Number of Students Eligible for Free/Reduced Lunch")
# can't see a great relationship, gonna remove the outliers to see a clearer relationship
```


```{r}
mutated <- schools_poverty %>%
  group_by(STABR) %>%
  summarise(
    "average IPR of each state" = mean(IPR_EST, na.rm = TRUE),
    "state" = tolower(abbr2state(STABR)),
    "lat" = mean(LAT, na.rm = TRUE),
    "lon" = mean(LON, na.rm = TRUE)
  )  
```

```{r}
ggplot(data = mutated) + 
  geom_map(
    aes(map_id = state, fill = `average IPR of each state`), 
    map = states) +
  expand_limits(x = states$long, y = states$lat) +
  coord_map() +
  theme_map() +
  labs(title = "Average Income Poverty Ratio for Each State") +
  theme(legend.position = "bottom")
```


\textbf{5.} Does the male to female ratio of students correlate with a higher or lower school neighborhood poverty estimate?

Does a higher school neighborhood poverty estimate correlate with a higher or lower free lunch program at schools?

```{r}
ggplot(schools_poverty, aes(x = IPR_EST, y = TOTFRL)) +
  geom_point() +
  geom_smooth(method = lm, se = FALSE) +
  labs(y = "Total number of students eligable
       for free and reduced lunch", x = "Income to poverty ratio, where lower IPR means higher poverty level", title = "Income to poverty ratio (IPR) and total students eligable for 
       reduced/free lunch at schools in the US")
```

```{r}
#making an error contour map
contour <- ggplot(states, aes(x=long, y=lat, group=group)) +  
  geom_polygon(color="navy", fill="lightblue") + 
  coord_fixed() +
  theme_map() +
  scale_color_viridis_d() +
  geom_density_2d(data = schools_poverty, aes(x = LON, y = LAT, group = FALSE, 
                                              colour = "IPR_EST"), alpha = 0.8, show.legend = TRUE) +
  labs(title = "Contour map of the U.S., showing levels IPR")

ggplotly(contour)
```

# Making full U.S. maps with all schools

```{r}
# Remove Hawaii and Alaska to zoom into the main 48 states
schools_48 <- schools_poverty %>%
  filter(STABR != "AK" & STABR != "HI") %>%
  mutate(
    "Income_Poverty_Ratio" = IPR_EST,
    "Student_Teacher_Ratio" = STUTERATIO, 
    "Black_Student_Ratio" = BL/TOTAL,
    "Male_To_Female_Ratio" = (TOTMENROL+1)/(TOTFENROL+1),
    "Ratio_of_Free_Lunches" = (TOTFRL+1)/(TOTAL+1)
  ) %>% select(Income_Poverty_Ratio, Student_Teacher_Ratio, Black_Student_Ratio, Male_To_Female_Ratio, Ratio_of_Free_Lunches, LAT, LON)

# Plot all of our variables: 

# IPR
ggplot(states, aes(x=long, y=lat, group=group)) +  
  geom_polygon(color="black", fill="white") + 
  coord_fixed() +
  theme_map() +
  geom_point(data = schools_48, aes(x = LON, y = LAT, group = FALSE, colour = Income_Poverty_Ratio), alpha = 0.5, shape = ".") +
  expand_limits(x = states$long, y = states$lat) +
  scale_color_viridis_c() +
  theme(legend.position = "bottom") +
  labs(title = "Income Poverty Ratio for Public Schools in U.S.", color = "IPR")

# Student Teacher Ratio
ggplot(states, aes(x=long, y=lat, group=group)) +  
  geom_polygon(color="black", fill="white") + 
  coord_fixed() +
  theme_map() +
  geom_point(data = schools_48, aes(x = LON, y = LAT, group = FALSE, colour = Student_Teacher_Ratio), alpha = 0.5, shape = ".") +
  expand_limits(x = states$long, y = states$lat) +
  scale_color_viridis_c() +
  theme(legend.position = "bottom") +
  labs(title = "Student Teacher Ratio for Public Schools in U.S.", color = "Student Teacher Ratio")

# Black Student Ratio
ggplot(states, aes(x=long, y=lat, group=group)) +  
  geom_polygon(color="black", fill="white") + 
  coord_fixed() +
  theme_map() +
  geom_point(data = schools_48, aes(x = LON, y = LAT, group = FALSE, colour = Black_Student_Ratio), alpha = 0.5, shape = ".") +
  expand_limits(x = states$long, y = states$lat) +
  scale_color_viridis_c() +
  theme(legend.position = "bottom") +
  labs(title = "Black Student Ratio for Public Schools in U.S.", color = "Black Student Ratio")

# Male to Female Ratio
ggplot(states, aes(x=long, y=lat, group=group)) +  
  geom_polygon(color="black", fill="white") + 
  coord_fixed() +
  theme_map() +
  geom_point(data = schools_48, aes(x = LON, y = LAT, group = FALSE, colour = Male_To_Female_Ratio), alpha = 0.5, shape = ".") +
  expand_limits(x = states$long, y = states$lat) +
  scale_color_viridis_c() +
  theme(legend.position = "bottom") +
  labs(title = "Male to Female Student Ratio for Public Schools in U.S.", color = "Male to Female Student Ratio")

# Free Lunch Program
ggplot(states, aes(x=long, y=lat, group=group)) +  
  geom_polygon(color="black", fill="white") + 
  coord_fixed() +
  theme_map() +
  geom_point(data = schools_48, aes(x = LON, y = LAT, group = FALSE, colour = Ratio_of_Free_Lunches), alpha = 0.5, shape = ".") +
  expand_limits(x = states$long, y = states$lat) +
  scale_color_viridis_c() +
  theme(legend.position = "bottom") +
  labs(title = "Ratio of Students using Free/Lunch Programs for Public Schools in U.S.", color = "Ratio of Students using Free/Lunch Programs")
```

# Leaflet

```{r}
colors_student_ratio <- colorNumeric(
  palette = c("red", "blue"),
  domain = range(schools_poverty$STUTERATIO, na.rm = TRUE)
)

leaflet(data = schools_poverty) %>%
  addTiles() %>%
  setView(lng = -98, lat = 40, zoom = 4) %>%
  addCircles(radius = 1, popup = paste("School: ", schools_poverty$SCH_NAME, ", Student to Teacher Ratio: ", schools_poverty$STUTERATIO), color = ~colors_student_ratio(schools_poverty$STUTERATIO))
```


```{r}
colors_student_ratio <- colorNumeric(
  palette = c("red", "blue"),
  domain = range(schools_poverty$Ratio_of_Free_Lunches, na.rm = TRUE)
)

leaflet(data = schools_poverty) %>%
  addTiles() %>%
  setView(lng = -98, lat = 40, zoom = 4) %>%
  addCircles(radius = 1, popup = paste("School: ", schools_poverty$SCH_NAME, ", Free/Reduced Lunch Ratio: ", schools_poverty$Ratio_of_Free_Lunches), color = ~colors_student_ratio(schools_poverty$Ratio_of_Free_Lunches))
```

